---
title: "Final Paper"
author: "Orsos Mihaly"
date: "2016. feb. 10."
output: html_document
---

Data analysis project bikeshare
===============================

```{r echo=FALSE, message=FALSE, include=FALSE, cache=FALSE, warning=FALSE}

library(dplyr)
library(ggplot2)
library(randomForest)
library(h2o)
library(pander)


```


Exploratory Analysis
--------------------

### Read data, quick look

```{r}

setwd("C:/Users/Mihály/Dropbox/Winter semester/Data Science for Business/my_code/bikeshare")
d <- read.csv("hour.csv", header = T, sep=",", stringsAsFactors = F)
dim(d)
head(d)
str(d)
colnames(d)
summary(d)
```


	- instant: record index
	- dteday : date
	- season : season (1:springer, 2:summer, 3:fall, 4:winter)
	- yr : year (0: 2011, 1:2012)
	- mnth : month ( 1 to 12)
	- hr : hour (0 to 23)
	- holiday : weather day is holiday or not (extracted from http://dchr.dc.gov/page/holiday-schedule)
	- weekday : day of the week
	- workingday : if day is neither weekend nor holiday is 1, otherwise is 0.
	+ weathersit : 
		- 1: Clear, Few clouds, Partly cloudy, Partly cloudy
		- 2: Mist + Cloudy, Mist + Broken clouds, Mist + Few clouds, Mist
		- 3: Light Snow, Light Rain + Thunderstorm + Scattered clouds, Light Rain + Scattered clouds
		- 4: Heavy Rain + Ice Pallets + Thunderstorm + Mist, Snow + Fog
	- temp : Normalized temperature in Celsius. The values are divided to 41 (max)
	- atemp: Normalized feeling temperature in Celsius. The values are divided to 50 (max)
	- hum: Normalized humidity. The values are divided to 100 (max)
	- windspeed: Normalized wind speed. The values are divided to 67 (max)
	- casual: count of casual users
	- registered: count of registered users
	- cnt: count of total rental bikes including both casual and registered
	
